### 2013-06-01 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding to spn estimator analysis section

\subsubsection{Alternative Fixed Point Iterations}
\label{subsubsec:alternative_fixed_point}

>>>>

\subsubsection{Alternative Fixed Point Iterations}
\label{subsubsec:alternative_fixed_point}
In addition to the basic Richardson iteration, the MCSA algorithm
presented in Eq~(\ref{eq:mcsa}) can be used to accelerate any fixed
point iteration which only depends on the state (typically the
residual) of the last iteration. As outlined in
Appendix~\ref{ch:linear_problem}, subspace methods take on a general
form in one-dimension using the Petrov-Galerkin conditions as
constraints for extracting a correction from the search subspace as
given by Eq~(\ref{eq:linear_projection_iteration}) that may be used
with MCSA. As an example, for positive-definite but not necessarily
symmetric problems, the minimal residual (MINRES) iteration
\citep{saad_iterative_2003} can be used in conjuction with MCSA where
the residual vector, $\mathbf{r}$, defines the search subspace and the
action of the linear operator on the residual vector,
$\mathbf{A}\mathbf{r}$, defines the constraint subspace. Using this,
we can then define an MCSA scheme that accelerates the MINRES
iteration:
\begin{subequations}
  \begin{gather}
    \alpha = \frac{\langle \mathbf{A}\mathbf{r}^k, \mathbf{r}^k
      \rangle}{\langle \mathbf{A}\mathbf{r}^k, \mathbf{A}\mathbf{r}^k
      \rangle}\:, \\
    \ve{x}^{k+1/2} = \ve{x}^k + \alpha \ve{r}^k\:,
    \label{eq:minres_correction}\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\ 
    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\ 
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta\ve{x}^{k+1/2}\:,\\ 
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:.
  \end{gather}
  \label{eq:mcsa_minres}
\end{subequations}
where $\alpha$ is an optimal extrapolation parameter generated the
constraints. Interestingly, this scheme is nearly identical to the
Richardson iteration version in Eq~(\ref{eq:mcsa}) except now the
additional extrapolation parameter, $\alpha$, is applied to the
residual correction in Eq~(\ref{eq:minres_correction}) as a means of
selecting a more optimal search direction at each iteration,
potentially further accelerating convergence.

### 2013-06-01 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding to spn estimator analysis section

\chapter{Monte Carlo Solution Methods Applied to Nonlinear Systems}
\label{ch:nonlinear_problem}
Nonlinear equation sets are a common occurrence in multiphysics
problems. Systems of partial differential equations such as those that
describe fluid flow or more general transport processes when
discretized by conventional methods yield discrete sets of stiff
equations with nonlinearities present in the variables. Traditionally,
such systems have been solved by linearizing them in a form where the
nonlinearities in the variables are eliminated and more traditional
linear methods can be used for solutions. Often characterized as
segregated methods where physics operators are split and their action
on the system approximated in steps, such methods lack consistency and
accuracy in resolving the nonlinear component of the solution. In the
last 30 years, fully consistent nonlinear methods based on Newton's
method have become more popular and many advances have been made in
the computational physics field to employ these methods.

>>>>

\chapter{Monte Carlo Solution Methods Applied to the Navier-Stokes Equations}
\label{ch:nonlinear_problem}
Nonlinear transport problems are a common occurrence in single and
multiple physics problems. Systems of partial differential equations
such as those that describe fluid flow or more general transport
processes when discretized by conventional methods yield discrete sets
of stiff equations with nonlinearities present in the
variables. Traditionally, such systems have been solved by linearizing
them in a form where the nonlinearities in the variables are
eliminated and more traditional linear methods can be used for
solutions. Often characterized as segregated methods where physics
operators are split and their action on the system approximated in
steps, such methods lack consistency and accuracy in resolving the
nonlinear component of the solution. In the last 30 years, fully
consistent nonlinear methods based on Newton's method have become more
popular and many advances have been made in the computational physics
field to employ these methods.

### 2013-06-01 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding to spn estimator analysis section

\subsection{Monte Carlo Estimator Comparison}
\label{subsec:spn_estimator_comparison}
With convergence obtained for the fuel assembly problem, the advanced
preconditioners will next be studied with both the collision and
expected value estimators using the best preconditioning and
relaxation parameter combinations found by the previous
analysis. Although the expected value estimator has achieved the best
iterative performance in the previous comparison and analysis for the
simple Poisson problem in \S~\ref{sec:parameter_estimator_analysis},
it is possible that the collision estimator may perform better from a
timing perspective due to the sparsity lost by explicit
preconditioning. The deterministic component of the expected value
estimator that couples the current state to other local states through
the iteration matrix stencil requires application of the estimator to
potentially several orders of magnitude more states at every
transition event.

For this study, the fuel assemby problem with a 1-group $SP_1$
discretization was solved using MCSA with the adjoint Monte Carlo
solver using both the collision and expected value estimators. Using
the results from the previous section, a Neumann relaxation parameter
of 0.7 was used with both estimators. For the collision estimator, a
Richardson relaxation parameter of 1.1 was used while it was found the
expected value estimator had the best performance when a a relaxation
parameter of 1.0 was used instead. In addition, ILUT preconditioning
with a drop tolerance of \sn{1}{-5} and a fill level of 5 was used
with a reduced domain approximation fill level of 100 and a threshold
of \sn{1}{-10} as in the previous analysis. For each estimator, the
number of stochastic histories used to compute the MCSA correction was
varied from 25 to 50,000 with the number of MCSA iterations required
to converge to a tolerance of \sn{1}{-8} recorded for a single
eigenvalue iteration along with the CPU time required to achieve
convergence.

Figure~\ref{fig:spn_estimator_iters} gives the iteration count results
and Figure~\ref{fig:spn_estimator_iters} gives the timing
results. Remarkably, although the fuel assembly problem is
significantly more complicated than the simple Poisson problem studied
in \S~\ref{sec:parameter_estimator_analysis}, the results here are
largely the same with Figure~\ref{fig:spn_estimator_iters} and
Figure~\ref{fig:estimator_nh_iters} having identical qualitative
behavior. Using the expected value estimator creates a relative
insensitivity to the number of histories used to compute the
correction. With respect to CPU time, there is a more pronounced
effect due to the larger density of states created by the explicit
ILUT preconditioning. Using a reduced domain fill level of 100 creates
a situation where the expected value estimator is significantly more
expensive per history to compute that the collision estimator due to
the coupling of states.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/estimator_iters.pdf}
  \end{center}
  \caption{\textbf{MCSA iterations required to converge the fuel
      assembly problem to a tolerance of \sn{1}{-8}.} \textit{Both the
      collision and expected value estimators were used with the
      adjoint Monte Carlo solver to compute the correction at each
      iteration.}}
  \label{fig:spn_estimator_iters}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/estimator_time.pdf}
  \end{center}
  \caption{\textbf{Total CPU time in seconds required to converge the
      fuel assembly problem to a tolerance of \sn{1}{-8}.}
    \textit{Both the collision and expected value estimators were used
      with the adjoint Monte Carlo solver to compute the correction at
      each iteration.}}
  \label{fig:spn_estimator_time}
\end{figure}

From both Figure~\ref{fig:spn_estimator_iters} and
Figure~\ref{fig:spn_estimator_iters} we see that both estimators have
comparable iterative and timing performance when 1,000 histories are
used at each MCSA iteration. Using 1,000 histories per iteration, the
infinity norm of the residual vector is presented at every MCSA
iteration for both estimators in
Figure~\ref{fig:spn_estimator_convergence} as an additional means of
comparison. At this number of histories, both estimators converge the
fuel assembly problem monotonically at and approximately the same
rate, giving little reason to choose one over the other. The end
result is that both estimators provide a viable method for getting
good MCSA iterative performance with reasonable timing
performance. For the collision estimator, using around 20,000
stochastic histories at every iteration gave the best iterative
performance while 500 histories per iteration gave the best iterative
performance when using the expected value estimator. For purely serial
computational performance, there is not a distinguishing feature for
the fuel assembly problem presented that would cause one to choose one
of the estimators over the other.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/estimator_convergence.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm as a function of MCSA
      iteration with 1,000 stochastic histories per iteration.}
    \textit{Both the collision and expected value estimators were used
      with the adjoint Monte Carlo solver to compute the correction at
      each iteration.}}
  \label{fig:spn_estimator_convergence}
\end{figure}

>>>>

\subsection{Monte Carlo Estimator Comparison}
\label{subsec:spn_estimator_comparison}
With convergence obtained for the fuel assembly problem, the advanced
preconditioners will next be studied with both the collision and
expected value estimators using the best preconditioning and
relaxation parameter combinations found by the previous
analysis. Additionally, both the Richardson-based MCSA iteration and
the MINRES-based MCSA iteration given by Eq~(\ref{eq:mcsa_minres})
will be used. Although the expected value estimator has achieved the
best iterative performance in the previous comparison and analysis for
the simple Poisson problem in
\S~\ref{sec:parameter_estimator_analysis}, it is possible that the
collision estimator may perform better from a timing perspective due
to the sparsity lost by explicit preconditioning. The deterministic
component of the expected value estimator that couples the current
state to other local states through the iteration matrix stencil
requires application of the estimator to potentially several orders of
magnitude more states at every transition event.

For this study, the fuel assemby problem with a 1-group $SP_1$
discretization was solved using MCSA with the adjoint Monte Carlo
solver using both the collision and expected value estimators and both
the Richardson and MINRES fixed point iterations. Using the results
from the previous section, a Neumann relaxation parameter of 0.7 was
used with both estimators and fixed point iterations. For the
collision estimator, a Richardson relaxation parameter of 1.1 was used
with the Richardson iteration while it was found the expected value
estimator had the best performance when a a relaxation parameter of
1.0 was used instead. In addition, ILUT preconditioning with a drop
tolerance of \sn{1}{-5} and a fill level of 5 was used with a reduced
domain approximation fill level of 100 and a threshold of \sn{1}{-10}
as in the previous analysis. For each estimator, the number of
stochastic histories used to compute the MCSA correction was varied
from 25 to 50,000 with the number of MCSA iterations required to
converge to a tolerance of \sn{1}{-8} recorded for a single eigenvalue
iteration along with the CPU time required to achieve convergence.

Figure~\ref{fig:spn_estimator_iters} gives the iteration count results
and Figure~\ref{fig:spn_estimator_time} gives the timing
results. Remarkably, although the fuel assembly problem is
significantly more complicated than the simple Poisson problem studied
in \S~\ref{sec:parameter_estimator_analysis}, the results here are
largely the same with Figure~\ref{fig:spn_estimator_iters} and
Figure~\ref{fig:estimator_nh_iters} having identical qualitative
behavior. Using the expected value estimator with both fixed point
iterations creates a relative insensitivity to the number of histories
used to compute the correction. For the collision estimator, the
MINRES iteration permits fewer stochastic histories to be used with
each MCSA iteration while still maintaining good iterative
performance. With respect to CPU time, there is a more pronounced
effect due to the larger density of states created by the explicit
ILUT preconditioning. Using a reduced domain fill level of 100 creates
a situation where the expected value estimator is significantly more
expensive per history to compute that the collision estimator due to
the coupling of states.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/estimator_iters.pdf}
  \end{center}
  \caption{\textbf{MCSA iterations required to converge the fuel
      assembly problem to a tolerance of \sn{1}{-8}.} \textit{Both the
      collision and expected value estimators were used with the
      adjoint Monte Carlo solver to compute the correction at each
      iteration. Each estimator was used with the Richardson and
      MINRES fixed point iterations within an MCSA iteration.}}
  \label{fig:spn_estimator_iters}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/estimator_time.pdf}
  \end{center}
  \caption{\textbf{Total CPU time in seconds required to converge the
      fuel assembly problem to a tolerance of \sn{1}{-8}.}
    \textit{Both the collision and expected value estimators were used
      with the adjoint Monte Carlo solver to compute the correction at
      each iteration. Each estimator was used with the Richardson and
      MINRES fixed point iterations within an MCSA iteration.}}
  \label{fig:spn_estimator_time}
\end{figure}

From both Figure~\ref{fig:spn_estimator_iters} and
Figure~\ref{fig:spn_estimator_time} we see that both estimators have
comparable iterative and timing performance when 1,000 histories are
used at each MCSA iteration. Using 1,000 histories per iteration, the
infinity norm of the residual vector is presented at every MCSA
iteration for both estimators in
Figure~\ref{fig:spn_estimator_convergence} as an additional means of
comparison. At this number of histories, both estimators converge the
fuel assembly problem monotonically at and approximately the same
rate, giving little reason to choose one over the other. The end
result is that both estimators provide a viable method for getting
good MCSA iterative performance with reasonable timing
performance. For the collision estimator, using around 20,000
stochastic histories at every iteration gave the best iterative
performance while 500 histories per iteration gave the best iterative
performance when using the expected value estimator. For purely serial
computational performance, there is not a distinguishing feature for
the fuel assembly problem presented that would cause one to choose one
of the estimators over the other.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/estimator_convergence.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm as a function of MCSA
      iteration with 1,000 stochastic histories per iteration.}
    \textit{Both the collision and expected value estimators were used
      with the adjoint Monte Carlo solver to compute the correction at
      each iteration. Each estimator was used with the Richardson and
      MINRES fixed point iterations within an MCSA iteration.}}
  \label{fig:spn_estimator_convergence}
\end{figure}
